\PYGEOMETRY{}: a Python library for the creation of Monte Carlo radiation transport physical geometries

### 2020-09-04 public/paper/pyg4ometry.tex

\begin{algorithm}[h]
  \SetKwFunction{Intersect}{Intersect}
  \SetAlgoLined
  \KwIn{Logical volume mesh $m$ with daughter volume meshes $d\in D$.}
\KwResult{Set, $S$, of overlapping pairs of meshes.}
$V \longleftarrow \emptyset$\;
$S \longleftarrow \emptyset$\;
\For{$d_1 \in D$}{
  \If{\Intersect{$d_1, m$}}{
    $S \longleftarrow S \cup \{(d_1, m)\}$\;
\For{$d_2 \in D$}{
 \If{$d_1 = d_2$ {\bf or} $(d_2, d_1) \in V$}{
  continue\;
}
\If{\Intersect{$d_1, d_2$}}{
  $S \longleftarrow S \cup \{(d_1, d_2)\}$\;
}
$V \longleftarrow V \cup \{(d_1, d_2)\}$\;
}
}
}
\caption{The overlap checking algorithm employed in \PYGEOMETRY{}.}
\end{algorithm}

Overlap detection in \PYGEOMETRY{} relies of the meshes generated for each solid, 
any detection algorithm will be approximate. Generally the overlap detection 
algorithm proceeds as 
\begin{enumerate}
\item Check if any of the daughter solid meshes have a non-null intersection with the compliment of mother solid mesh
\item Loop over all daughter volumes and check
\item Check
\end{enumerate}

@@

\subsection{Materials}

Any useful translation between geometry description formats must also
account for material, and accordingly \PYGEOMETRY{} correctly translates
FLUKA materials to GDML.  FLUKA materials can be divided into built-in ,
single-element and compound materials.  Built-in materials are simply those
that are predefined by FLUKA, single-element materials are described with a
single \fluka{material} card, and compound materials are described with one
\fluka{material} card followed by one or more \fluka{compound} cards.
These three alternatives are represented in \PYGEOMETRY{} with the
\pyinline{BuiltIn}, \pyinline{Material}, and \pyinline{Compound} classes.
Populating a hierarchy of instances using these classes is involved due to
the fact that recursively-defined materials in FLUKA input files need not
be defined in a logical order.  Namely, a given \pyinline{compound} may be
defined before the materials that it consists of are themselves defined.
To account for this it is necessary to correctly compute the instantiation
order so that the above classes and be instantiated correctly.  To do this
a directed acyclic graph is populated with the materials and their
constituents, after which a topological sort is performed so that the
compound materials are sequenced after their constituents.  Mapping this
set of nested FLUKA material instances to GDML material instances is then
straight forwards as the GDML material semantics are slightly more
expressive than FLUKA's, so a one-to-one mapping is trivial.

@@

Figure~\ref{fig:fluka-to-geant4-cup} shows a Faraday cup implemented in
FLUKA and accurately translated to GDML using \PYGEOMETRY{}.  Many of the
steps described above were directly applied in this model and all the
features are tested and demonstrated in the repository.  This set of
algorithms for bridging FLUKA with GDML covers a very broad range of
geometries, however a number of possible improvements for the future
persist.  For example, Boolean solids in FLUKA can in general be
disconnected, and this most typically manifests itself in the form of
disconnected unions, but is also allowed for intersections and
subtractions.  Disconnected unions are readily availabe in GDML with the
use of the MultiUnion solid type, however there is no valid means to
construct a disconnected subtraction or intersection in GDML.  One possible
solution would be to detect these two cases, and if found, split them into
their constituent parts and place separately as TesselatedSolid instances.

@@

\section{GDML to FLUKA conversion}
It is relatively straight forward to convert Geant4 geometry to FLUKA. Each of the Geant4 
solids can be mapped to a FLUKA \verb|REGION|. Some solids in Geant4 directly map to 
a FLUKA body, others need a construction of a simple FLUKA CSG tree. Table~\ref{tab:geant2fluka}
lists the mapping between Geant4 solids and the bodies used to compose a FLUKA region.
A region is a region of space which is formed of bodies and the operators + (intersection), - 
(complement) and | (union), also parentheses to indicate the order of operations. FLUKA 
has 20 bodies, listed in Table~\ref{tab:Fluka2Geant4}  and in general infinite extent bodies have benefits over finite ones. Key 
for conversion are XY-, XZ-, YZ-Planes (XYP, XZP, YZP), arbitrary plane (PLA), 
Z axis aligned cylinder (ZCC), Z axis aligned elliptical cylinder (ZEC), 
sphere (SPH), truncated right-angle cone (TRC) and general quadric surface (QUA).

@@

Figure~\ref{fig:gdml-flair} shows an example conversion from GDML to FLUKA. The example is a 
vacuum chamber with three ConFlat flange (CF) beam pipes connected to CF flanges. The top and bottom plates 
have been removed to display the geometry more clearly. The model is formed of \verb|G4Box| and
\verb|G4Tubs| and Boolean operations of subtraction, intersection and union.
\begin{figure*}
\begin{center}
\includegraphics[width=1.0\textwidth]{./model-scene/CuboidalChamber.pdf}
\caption{Example conversion of a simple GDML geometry to FLUKA, left panel:  
the original model in \PYGEOMETRY{}, right panel: the converted FLUKA geometry 
viewed in FLAIR. The example is a sector bend dipole electromagnet.}
\label{fig:gdml-flair}
\end{center}
\end{figure*}

@@

\section{CAD/STL to GDML conversion }
STEP and IGES files can be loaded into \PYGEOMETRY{}, via an interface based on FreeCAD. FreeCAD is an 
open source CAD/CAM program, which in turn is based on OpenCASCADE. A STEP CAD file could be 
considered as a hierarchical tree of {\it part assemblies} and {\it part features}, where a part assembly is a 
collection of part features. A part feature can be used to create  a triangular mesh which can used to create 
a \PYGEOMETRY{} tessellated solid. Assignment of materials and visualisation attributes must be performed by 
the user after conversion to GDML as rarely will CAD/CAM packages included the detailed information required 
for RT codes.
\begin{lstlisting}[caption={A simple \PYGEOMETRY{} Python script to load a STEP file.},label={lst:pythonCADLoading}, language=Python]
    reader = pyg4ometry.freecad.Reader("CadFileName.step")
    reader.relabelModel()
    reader.convertFlat()
    logical = reader.getRegistry().getWorldVolume()
\end{lstlisting}

\begin{figure*}
\begin{center}
\includegraphics[width=1.0\textwidth]{./model-scene/SectorBend.pdf}
\caption{Example conversion of a simple FLUKA geometry to GDML, left panel:  
the original model in FreeCAD, right panel: the GDML geometry 
viewed in \PYGEOMETRY{}. The example is a sector bend dipole electromagnet.}
\label{fig:cad-gdml}
\end{center}
\end{figure*}

\section{Complete simulation example}
\PYGEOMETRY{} is designed to be as flexible as possible and offer the user a wide range of usage styles 
and workflows. The following sections describe a selection of 
possible applications.
%
\begin{enumerate}
\item Merging Registries 
\item Removing volumes 
\item Editing solid
\item Converting to assembly solids 
\end{enumerate}
%
\begin{figure*}[htb!]
\begin{center}
\includegraphics[width=1.0\textwidth]{./model-scene/model.pdf}
\caption{Complete compositing example using \PYGEOMETRY{}.}
\label{fig:model}
\end{center}
\end{figure*}

>>>>

\begin{algorithm}[h]
  \SetKwFunction{Intersect}{Intersect}
  \SetAlgoLined
  \KwIn{Logical volume mesh $m$ with daughter volume meshes $d\in D$.}
\KwResult{Set, $S$, of overlapping pairs of meshes.}
$V \longleftarrow \emptyset$\;
$S \longleftarrow \emptyset$\;
\For{$d_1 \in D$}{
  \If{\Intersect{$d_1, m$}}{
    $S \longleftarrow S \cup \{(d_1, m)\}$\;
\For{$d_2 \in D$}{
 \If{$d_1 = d_2$ {\bf or} $(d_1, d_2) \in V$}{
  continue\;
}
\If{\Intersect{$d_1, d_2$}}{
  $S \longleftarrow S \cup \{(d_1, d_2)\}$\;
}
$V \longleftarrow V \cup \{(d_1, d_2)\}$\;
}
}
}
\label{algo:overlap}
\caption{The overlap checking algorithm employed in \PYGEOMETRY{}.}
\end{algorithm}

Overlap detection in \PYGEOMETRY{} relies of the meshes generated for each solid, 
any detection algorithm will be approximate. Generally the overlap detection 
algorithm proceeds as shown in Algorithm~\ref{algo:overlap}.

@@

@@

Figure~\ref{fig:fluka-to-geant4-cup} shows a Faraday cup implemented in
FLUKA and translated to GDML using \PYGEOMETRY{}.  The conversion is clearly
accurate well and also includes the translation of the materials to GDML.
Many of the steps described above were directly applied in this model and
all the features are tested and demonstrated in the repository.  This set
of algorithms for bridging FLUKA with GDML covers a very broad range of
geometries, however a number of possible improvements for the future
persist.  For example, Boolean solids in FLUKA can in general be
disconnected, and this most typically manifests itself in the form of
disconnected unions, but is also allowed for intersections and
subtractions.  Disconnected unions are readily availabe in GDML with the
use of the MultiUnion solid type, however there is no valid means to
construct a disconnected subtraction or intersection in GDML.  One possible
solution would be to detect these two cases, and if found, split them into
their constituent parts and place separately as TesselatedSolid instances.

@@

\section{GDML to FLUKA conversion}
It is relatively straight forward to convert Geant4 geometry to FLUKA. Each of the Geant4 
solids can be mapped to a FLUKA \verb|REGION|. Some solids in Geant4 directly map to 
a FLUKA body, others need a construction of a simple FLUKA CSG tree. Table~\ref{tab:geant2fluka}
lists the mapping between Geant4 solids and the bodies used to compose a FLUKA region.
A region is a region of space which is formed of bodies and the operators + (intersection), - 
(complement) and | (union), also parentheses to indicate the order of operations. FLUKA 
has 20 bodies, listed in Table~\ref{tab:Fluka2Geant4}  and in general infinite extent bodies have 
benefits over finite ones. Key for conversion are XY-, XZ-, YZ-Planes (XYP, XZP, YZP), arbitrary 
plane (PLA), Z axis aligned cylinder (ZCC), Z axis aligned elliptical cylinder (ZEC), 
sphere (SPH), truncated right-angle cone (TRC) and general quadric surface (QUA). 
\begin{figure*}
\begin{center}
\includegraphics[width=1.0\textwidth]{./model-scene/CuboidalChamber.pdf}
\caption{Example conversion of a simple GDML geometry to FLUKA, left panel:  
the original model in \PYGEOMETRY{}, right panel: the converted FLUKA geometry 
viewed in FLAIR. The example is a sector bend dipole electromagnet.}
\label{fig:gdml-flair}
\end{center}
\end{figure*}

@@

Figure~\ref{fig:gdml-flair} shows an example conversion from GDML to FLUKA. The example is a 
vacuum chamber with three ConFlat flange (CF) beam pipes connected to CF flanges. The top and bottom plates 
have been removed to display the geometry more clearly. The model is formed of \verb|G4Box| and
\verb|G4Tubs| and Boolean operations of subtraction, intersection and union.

@@

\section{STL and CAD to GDML conversion}
STL and CAD conversion are closely related. In both cases the solid 
(in the case of STL) and solids (in the case of CAD) are converted to 
tessellated solid(s). STL is a relatively simple file format that can
be loaded using pure Python. As STL files typically only contain a single 
solid the \pyinline{stl.Reader} provides a single solid and not a logical volume
as with the other file readers.

STEP and IGES files can be loaded into 
\PYGEOMETRY{}, via an interface based on FreeCAD. FreeCAD is an 
open source CAD/CAM program, which in turn is based on OpenCASCADE. 
FreeCAD allows for scripting in Python and acts as a simple to use 
interface to OpenCASCADE.  A STEP CAD file could be considered as a 
hierarchical tree of part features and part assemblies, where a part assembly is a 
collection of part features. A part feature can be used to create a triangular mesh 
which can used to create a \PYGEOMETRY{} tessellated solid. The placement 
of the part feature is extracted from the STEP file and used to create an 
appropriate physical volume. Assignment of materials and visualisation 
attributes must be performed by  the user after conversion to GDML as rarely 
will CAD/CAM packages included the detailed information required for RTMC 
codes. Listing~\ref{lst:pythonCADLoading} shows 
how \PYGEOMETRY{} can load step files.
\begin{lstlisting}[caption={A simple \PYGEOMETRY{} Python script to load a STEP file.},label={lst:pythonCADLoading}, language=Python]
    reader = pyg4ometry.freecad.Reader("CadFileName.step")
    reader.relabelModel()
    reader.convertFlat()
    logical = reader.getRegistry().getWorldVolume()
\end{lstlisting}

An example geometry representing a dipole electromagnet, consisting of three parts was created in Autodesk Fusion 360
and saved as a STEP file, \PYGEOMETRY{} produces the output in Figure~\ref{fig:cad-gdml}
\begin{figure*}
\begin{center}
\includegraphics[width=1.0\textwidth]{./model-scene/SectorBend.pdf}
\caption{Example conversion of a simple CAD (STEP) geometry (sector bend dipole electromagnet) to GDML, left panel:  
the original model in FreeCAD, right panel: the GDML geometry 
viewed in \PYGEOMETRY{}. }
\label{fig:cad-gdml}
\end{center}
\end{figure*}

\section{Complete simulation example}
\PYGEOMETRY{} is designed to be as flexible as possible and offer the user a wide range of usage styles, input files 
and workflows. A fictitious beam line was created to demonstrate the capabilities of \PYGEOMETRY, this creates a composite 
{\it scene} which consists of the different examples presented in this paper.
%
\begin{figure*}
\begin{center}
\includegraphics[width=1.0\textwidth]{./model-scene/model.pdf}
\caption{Complete compositing example using \PYGEOMETRY{}.}
\label{fig:model}
\end{center}
\end{figure*}

\begin{enumerate}
\item Merging Registries 
\item Removing volumes 
\item Editing solid
\item Converting to assembly solids 
\end{enumerate}
%
